---
title: "REPORT_PHASE1 -- Semantic Fluency & Mental Lexicon"
author: "Brainiacs"
date: "`r Sys.Date()`"
output:
  pdf_document:
    toc: true
    toc_depth: 2
    number_sections: true
    fig_width: 7
    fig_height: 4
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                      fig.align = "center")
```

```{r load-and-prep, message=FALSE, warning=FALSE}
# -- Libraries ------------------------------------------------------------------
library(dplyr)
library(readr)
library(stringr)
library(ggplot2)
library(scales)
library(tidyr)

# -- Load Data ------------------------------------------------------------------
df_raw <- read_csv("final_dataset.csv")

# -- Filters & Feature Engineering ---------------------------------------------
df <- df_raw %>%
  mutate(
    state_ut = str_trim(state_ut),
    language  = str_to_lower(str_trim(language))
  ) %>%
  filter(language != "english") %>%
  mutate(
    Native_Status = ifelse(
      state_ut %in% c("Kerala", "Telangana", "Andhra Pradesh"),
      "Non-Native", "Native"
    ),
    Native_Status = factor(Native_Status, levels = c("Native", "Non-Native"))
  ) %>%
  group_by(subject_id, domain) %>%
  mutate(
    tag          = row_number(),
    IRT_ms       = latency_ms,
    rt           = cumsum(latency_ms),
    Prev_X       = lag(x_norm),
    Prev_Y       = lag(y_norm),
    Seq_Distance = sqrt((x_norm - Prev_X)^2 + (y_norm - Prev_Y)^2)
  ) %>%
  ungroup()

# Shared colour palette (colour-blind friendly)
pal <- c("Native" = "#2196F3", "Non-Native" = "#F44336")
```

------------------------------------------------------------------------

# H1 . Lexical Productivity & Access Speed

**Hypothesis:** Native Hindi speakers produce more unique words *and*
retrieve them faster (shorter IRT) than Non-Native speakers.

*Rationale:* A well-practised mental lexicon has denser, more
inter-connected nodes, making both quantity and speed superior for
native users.

```{r h1-data}
h1_data <- df %>%
  group_by(subject_id, Native_Status) %>%
  summarise(
    Unique_Words = n_distinct(word),
    Mean_IRT_s   = mean(IRT_ms, na.rm = TRUE) / 1000,
    .groups = "drop"
  )
```

```{r h1-plot-a, fig.width=6, fig.height=4}
ggplot(h1_data, aes(x = Native_Status, y = Unique_Words, fill = Native_Status)) +
  geom_violin(alpha = 0.35, trim = FALSE) +
  geom_boxplot(width = 0.25, outlier.shape = NA, alpha = 0.85) +
  geom_jitter(width = 0.1, size = 1.6, alpha = 0.55, colour = "grey30") +
  scale_fill_manual(values = pal) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "none") +
  labs(title = "H1A -- Total Unique Words per Participant",
       y = "Unique Words Generated", x = NULL)
```

```{r h1-plot-b, fig.width=6, fig.height=4}
ggplot(h1_data, aes(x = Native_Status, y = Mean_IRT_s, fill = Native_Status)) +
  geom_violin(alpha = 0.35, trim = FALSE) +
  geom_boxplot(width = 0.25, outlier.shape = NA, alpha = 0.85) +
  geom_jitter(width = 0.1, size = 1.6, alpha = 0.55, colour = "grey30") +
  scale_fill_manual(values = pal) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "none") +
  labs(title = "H1B -- Mean Inter-Response Time",
       y = "Mean IRT (seconds)", x = NULL)
```

```{r h1-plot-c, fig.width=7, fig.height=4}
ggplot(h1_data, aes(x = Unique_Words, y = Mean_IRT_s,
                    colour = Native_Status, shape = Native_Status)) +
  geom_point(size = 2.5, alpha = 0.7) +
  geom_smooth(method = "lm", se = TRUE, linewidth = 1) +
  scale_colour_manual(values = pal) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom") +
  labs(title = "H1C -- Productivity vs. Speed (per participant)",
       x = "Unique Words", y = "Mean IRT (s)",
       colour = NULL, shape = NULL)
```

### Observations


# H5 . Repeat / Intrusion Rate

**Hypothesis:** Non-Native speakers will show a higher rate of word
repetitions due to weaker lexical monitoring.

```{r h5-data}
h5_data <- df %>%
  group_by(subject_id, Native_Status) %>%
  summarise(
    Total_Words  = n(),
    Unique_Words = n_distinct(word),
    Repeat_Rate  = (Total_Words - Unique_Words) / Total_Words,
    .groups = "drop"
  )
```

```{r h5-plot-a, fig.width=6, fig.height=4}
ggplot(h5_data, aes(x = Native_Status, y = Repeat_Rate, fill = Native_Status)) +
  geom_violin(alpha = 0.35, trim = FALSE) +
  geom_boxplot(width = 0.25, outlier.shape = NA, alpha = 0.85) +
  geom_jitter(width = 0.1, size = 1.8, alpha = 0.55, colour = "grey30") +
  scale_y_continuous(labels = percent_format(accuracy = 1)) +
  scale_fill_manual(values = pal) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "none") +
  labs(title = "H5A -- Repeat Rate by Group",
       y = "% Words Repeated", x = NULL)
```

```{r h5-plot-b, fig.width=7, fig.height=4}
ggplot(h5_data, aes(x = Unique_Words, y = Repeat_Rate,
                    colour = Native_Status, shape = Native_Status)) +
  geom_point(size = 2.5, alpha = 0.7) +
  geom_smooth(method = "lm", se = TRUE, linewidth = 1) +
  scale_colour_manual(values = pal) +
  scale_y_continuous(labels = percent_format(accuracy = 1)) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom") +
  labs(title = "H5B -- Repeat Rate vs. Productivity",
       x = "Unique Words", y = "Repeat Rate (%)",
       colour = NULL, shape = NULL)
```

### Observations

-   **H5A:** A higher box for Non-Natives confirms weaker inhibitory
    control -- they re-enter already-produced words more often.

-   **H5B:** If both groups show a negative relationship (more words -\>
    fewer repeats), that is expected. A flat or positive slope for
    Non-Natives would indicate poor lexical self-monitoring even among
    productive speakers.

------------------------------------------------------------------------

# H6 . Domain-Specific Cognitive Load

**Hypothesis:** The IRT gap between groups will be *largest* for
abstract or culturally embedded domains (e.g., Furniture) and smallest
for universal, embodied categories (Body Parts).

```{r h6-data}
h6_data <- df %>%
  mutate(IRT_s  = IRT_ms / 1000,
         domain = str_to_title(str_replace_all(domain, "-", " ")))

h6_mean <- h6_data %>%
  group_by(domain, Native_Status) %>%
  summarise(Mean_IRT_s = mean(IRT_s, na.rm = TRUE), .groups = "drop")
```

```{r h6-plot-a, fig.width=8, fig.height=4}
ggplot(h6_data, aes(x = domain, y = IRT_s, fill = Native_Status)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.75) +
  coord_cartesian(ylim = c(0, 15)) +
  scale_fill_manual(values = pal) +
  theme_minimal(base_size = 11) +
  theme(axis.text.x = element_text(angle = 25, hjust = 1),
        legend.position = "bottom") +
  labs(title = "H6A -- IRT Distribution by Domain",
       y = "IRT (seconds)", x = NULL, fill = NULL)
```

```{r h6-plot-b, fig.width=6, fig.height=4}
ggplot(h6_mean, aes(x = Native_Status, y = domain, fill = Mean_IRT_s)) +
  geom_tile(colour = "white", linewidth = 0.8) +
  geom_text(aes(label = round(Mean_IRT_s, 1)), size = 3.8, fontface = "bold") +
  scale_fill_gradient(low = "#E3F2FD", high = "#B71C1C",
                      name = "Mean IRT (s)") +
  theme_minimal(base_size = 12) +
  theme(axis.text.y = element_text(face = "bold")) +
  labs(title = "H6B -- Mean IRT Heatmap",
       x = NULL, y = NULL)
```

### Observations

-   **H6A (Boxplots):** The vertical gap between box medians for each
    domain is the direct measure of additional cognitive load for
    Non-Natives.

-   **H6B (Heatmap):** Darker cells = higher mean IRT = harder domain.
    If Non-Native cells are consistently darker, the group difference is
    robust across domains.

------------------------------------------------------------------------

# H7 . Distance-Time Coupling

**Hypothesis:** For Native speakers, larger spatial jumps will *predict*
longer IRTs. Non-Natives will show a *weaker or decoupled* relationship.

```{r h7-plot, fig.width=9, fig.height=4}
h7_sub <- df %>%
  filter(!is.na(Seq_Distance), !is.na(IRT_ms)) %>%
  mutate(IRT_s = IRT_ms / 1000)

ggplot(h7_sub, aes(x = Seq_Distance, y = IRT_s, colour = Native_Status)) +
  geom_point(alpha = 0.12, size = 1.2) +
  geom_smooth(method = "lm", se = TRUE, linewidth = 1.5) +
  coord_cartesian(ylim = c(0, 20)) +
  scale_colour_manual(values = pal) +
  facet_wrap(~ str_to_title(str_replace_all(domain, "-", " ")), nrow = 1) +
  theme_minimal(base_size = 10) +
  theme(legend.position = "bottom",
        strip.text = element_text(face = "bold")) +
  labs(
    title = "H7 -- Spatial Distance-Time Coupling by Domain",
    subtitle = "Steeper slope = tighter link between space and retrieval time",
    x = "Sequential Spatial Distance", y = "IRT (seconds)", colour = NULL
  )
```

### Observations

-   A **steeper positive slope** for Native speakers means space and
    time are yoked -- evidence of coherent spatial encoding.

-   A **flat or near-zero slope** for Non-Natives indicates decoupling:
    spatial placements are not driven by retrieval order.

-   Domains like *Animals* (larger vocabulary) should show the strongest
    coupling; Furniture Practice may show flat slopes for both groups.

------------------------------------------------------------------------

# H8 . Boundary Effect -- The Exit Pause

**Hypothesis:** The IRT immediately *before* a large spatial jump will
be significantly longer than IRTs within a cluster, reflecting the
cognitive cost of switching mental patches.

```{r h8-data}
h8_data <- df %>%
  filter(!is.na(Seq_Distance), tag > 1) %>%
  mutate(
    Movement_Type = ifelse(Seq_Distance > 0.4,
                           "Cluster Switch (Large Jump)",
                           "Within-Cluster (Small Jump)"),
    IRT_s = IRT_ms / 1000
  )

h8_prop <- df %>%
  filter(!is.na(Seq_Distance), tag > 1) %>%
  group_by(subject_id, Native_Status) %>%
  summarise(
    Big_Jump_Prop = mean(Seq_Distance > 0.4, na.rm = TRUE),
    .groups = "drop"
  )
```

```{r h8-plot-a, fig.width=7, fig.height=4}
ggplot(h8_data, aes(x = Movement_Type, y = IRT_s, fill = Native_Status)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.8) +
  coord_cartesian(ylim = c(0, 15)) +
  scale_fill_manual(values = pal) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom") +
  labs(title = "H8A -- Exit Pause by Movement Type",
       x = NULL, y = "IRT (s)", fill = NULL)
```

```{r h8-plot-b, fig.width=6, fig.height=4}
ggplot(h8_prop, aes(x = Native_Status, y = Big_Jump_Prop, fill = Native_Status)) +
  geom_violin(alpha = 0.35, trim = FALSE) +
  geom_boxplot(width = 0.25, outlier.shape = NA, alpha = 0.85) +
  scale_y_continuous(labels = percent_format(accuracy = 1)) +
  scale_fill_manual(values = pal) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "none") +
  labs(title = "H8B -- Proportion of Large Spatial Jumps",
       x = NULL, y = "% Large Spatial Jumps")
```

### Observations

-   **H8A:** Cluster Switch boxes should sit higher than Within-Cluster
    boxes for both groups. The key comparison is how large this gap is
    for Natives vs. Non-Natives.

-   If Natives show a bigger pause before switching, they are more aware
    of cluster boundaries. Smaller or equal pauses for Non-Natives
    suggest their switches are less deliberate.

-   **H8B:** A higher proportion of big jumps for Non-Natives indicates
    a more fragmented, less locally clustered spatial strategy.

------------------------------------------------------------------------
